% arara: xelatex
\documentclass[12pt]{article}

\usepackage{tikz} % картинки в tikz
\usepackage{microtype} % свешивание пунктуации

\usepackage{array} % для столбцов фиксированной ширины

\usepackage{indentfirst} % отступ в первом параграфе

\usepackage{sectsty} % для центрирования названий частей
\allsectionsfont{\centering}

\usepackage{amsmath, amsthm, amssymb} % куча стандартных математических плюшек

\usepackage{comment}
\usepackage{amsfonts}

\usepackage[top=2cm, left=1cm, right=1cm, bottom=2cm]{geometry} % размер текста на странице

\usepackage{lastpage} % чтобы узнать номер последней страницы

\usepackage{enumitem} % дополнительные плюшки для списков
%  например \begin{enumerate}[resume] позволяет продолжить нумерацию в новом списке
\usepackage{caption}

\usepackage{hyperref} % гиперссылки

\usepackage{multicol} % текст в несколько столбцов


\usepackage{fancyhdr} % весёлые колонтитулы
\pagestyle{fancy}
\lhead{Stochastic calculus — ICEF}
\chead{2023}
\rhead{Home assignment}
\lfoot{}
\cfoot{}
\rfoot{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}



\usepackage{todonotes} % для вставки в документ заметок о том, что осталось сделать
% \todo{Здесь надо коэффициенты исправить}
% \missingfigure{Здесь будет Последний день Помпеи}
% \listoftodos — печатает все поставленные \todo'шки


% более красивые таблицы
\usepackage{booktabs}
% заповеди из докупентации:
% 1. Не используйте вертикальные линни
% 2. Не используйте двойные линии
% 3. Единицы измерения - в шапку таблицы
% 4. Не сокращайте .1 вместо 0.1
% 5. Повторяющееся значение повторяйте, а не говорите "то же"


\usepackage{fontspec}
\usepackage{polyglossia}

\setmainlanguage{english}
\setotherlanguages{russian}

% download "Linux Libertine" fonts:
% http://www.linuxlibertine.org/index.php?id=91&L=1
\setmainfont{Linux Libertine O} % or Helvetica, Arial, Cambria
% why do we need \newfontfamily:
% http://tex.stackexchange.com/questions/91507/
\newfontfamily{\cyrillicfonttt}{Linux Libertine O}

\AddEnumerateCounter{\asbuk}{\russian@alph}{щ} % для списков с русскими буквами
% \setlist[enumerate, 2]{label=\asbuk*),ref=\asbuk*}

%% эконометрические сокращения
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Corr}{Corr}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\E}{E}
\def \hb{\hat{\beta}}
\def \hs{\hat{\sigma}}
\def \htheta{\hat{\theta}}
\def \s{\sigma}
\def \hy{\hat{y}}
\def \hY{\hat{Y}}
\def \v1{\vec{1}}
\def \e{\varepsilon}
\def \he{\hat{\e}}
\def \z{z}
\def \hVar{\widehat{\Var}}
\def \hCorr{\widehat{\Corr}}
\def \hCov{\widehat{\Cov}}
\def \cN{\mathcal{N}}
\def \cF{\mathcal{F}}
\def \P{\mathbb{P}}


\begin{document}


\section*{Submission details}

% Submit you home assignments at \url{https://www.gradescope.com} with coursecode: NXE75N. 

Deadline HA1: 26 November 2023, 23:59

Deadline HA2: 07 December 2023, 23:59

Deadline HA3: 17 December 2023, 23:59

You have one honey-day. 
The honey-day allows you to postpone one of the three deadlines by 24 hours. 

\newpage



\section{HA-1}

\begin{enumerate}

\item 
  Consider the following joint distribution of $X$ and $Y$:

  \begin{tabular}{*{4}{c}}
  \toprule
  & $X=-1$ & $X=0$ & $X=1$ \\
  \midrule
  $Y=0$ & 0.1 & 0.2 & 0.3  \\
  $Y=1$ & 0.2 & 0.1 & 0.1  \\
  \bottomrule
  \end{tabular}
  
  \begin{enumerate}
   \item Find explicitely $\sigma(X)$, $\sigma(Y)$, $\sigma(X \cdot Y)$, $\sigma(X^2)$, $\sigma(2X+3)$.
   \item Home many elements are there in $\sigma(X, Y)$, $\sigma(X + Y)$, $\sigma(X, Y, X+Y)$?
  \end{enumerate}
  

\item More $\sigma$-algebra questions :)
\begin{enumerate}
  \item You observe the result of 10 independent coin tosses. How many elements the corresponding $\sigma$-algebra contains?
  \item Prove that a finite $\sigma$-algebra can contain only $2^k$ elements.
  \item Is union of two $\sigma$-algebras always a $\sigma$-algebra? Prove your statement.
  \item Is intersection of two $\sigma$-algebras always a $\sigma$-algebra? Prove your statement.
\end{enumerate}

\item 
Is it true that for any two $\sigma$-algebras $\mathcal{F}$ and $\mathcal{H}$ and for any random variable $Y$
\[
  \E(\E(Y|\mathcal F)|\mathcal H) = \E(Y|\mathcal F \cap \mathcal H)?
\]
Prove the statement or provide a counter-example.
\item I throw a fair die until the first six appears. 
Let's denote the total number of throws by $X$ and the number of odd integers thrown by $Y$.

\begin{enumerate}
  \item Find $\P(Y=y|X)$, $\E(Y|X)$, $\Var(Y|X)$;
  \item Find $\E(X|Y)$.
\end{enumerate}

\item I throw 100 coins. Let's denote by $X$ the number of coins that show «heads».
I throw these $X$ coins once again, leaving other coins as they are. 
Let's denote by $Y$ the number of coins that show «heads» now.

Find $\P(Y=y|X)$, $\E(Y|X)$, $\Var(Y|X)$, $\E(Y)$, $\Var(Y)$.


% \item Let $X\sim U[0;1]$, and $Y$ takes values 0 and 1 with equal probabilities. 
% Random variables $X$ and $Y$ are independent and $Z=X^{Y}$.

% Find $\E(Z|Y)$, $\Var(Z|Y)$, $\E(Z|X)$, $\Var(Z|X)$.

\item Random variables $X$ and $Y$ have joint normal distribution with zero means and covariance matrix
\[
\begin{pmatrix}
4 & -1 \\
-1 & 9 \\
\end{pmatrix}.
\]

\begin{enumerate}
  \item Find $\E(Y|X)$, $\Var(Y|X)$, $\E(XY|X)$ and $\Var(XY|X)$.
  \item Using standard normal cumulative distribution function find $\P(YX > 2021 | X)$.
\end{enumerate}



\end{enumerate}

\newpage
\section{HA-2}

\begin{enumerate}

\item The random variables $Z_1$, $Z_2$, \ldots{} are independent and identically distributed 
with $\P(Z_n = 1) = 0.7$ and $\P(Z_n = -1) = 0.3$. 
Consider the cumulative sum process, $S_n = Z_1 + \ldots + Z_n$ with $S_0=0$.

\begin{enumerate}
  \item Find all values of $a$ such that $\exp(aS_n)$ is a martingale.
  \item If possible find the constants $\alpha$ and $\beta$ such that $Y_n = S_n^2 + \alpha S_n + \beta n$ is a martingale.
\end{enumerate}

\item The random variables $Z_1$, $Z_2$, \ldots{} are independent and identically distributed 
with $\P(Z_n = +1) = 0.1$,  $\P(Z_n = -1) = 0.1$ and $\P(Z_n = 0) = 0.8$. 
Consider the cumulative sum process, $S_n = Z_1 + \ldots + Z_n$ with $S_0=0$.

Let $\tau$ be the first moment when $S_n = 10$ or $S_n = -20$.

\begin{enumerate}
  \item Is $S_n$ a martingale?
  \item Find $\P(S_{\tau} = 10)$.
  \item If possible find the non-random function $a_n$ such that $Y_n = S_n^2 + a_n$ is a martingale. 
  \item Find $\E(\tau)$.
\end{enumerate}

\item Let $(W_t)$ be a standard Wiener process.
\begin{enumerate}
  \item Find $\E(\sin(\alpha W_t))$.
  \item Find $\E(\exp(\alpha W_t))$.
  \item Find $\E(\cos(\alpha W_t))$.
\end{enumerate}
Hint: you may solve this with or without Ito's lemma, that's up to you. 

\item Let $(W_t)$ be a standard Wiener process and $Y_t = W_t^3 + t^2 W_t^2$.
\begin{enumerate}
  \item Find $\E(Y_t)$ and $\Var(Y_t)$.
  \item Is $Y_t$ a martingale?
  \item Find $\E(Y_{t} \mid W_s)$ for $t \geq s$. 
\end{enumerate}

\item The processes $(X_t)$ and $(Y_t)$ are independent Wiener processes adapted to the filtration $(\mathcal{F}_t)$.
The process $Z_t = aX_t + 0.3 Y_t$ with $a>0$ is also a Wiener process. 
\begin{enumerate}
  \item Find the value of $a$.
  \item Find the correlation $\Corr(Z_t, X_t)$.
  \item Find $\E(Z_3 \mid X_2)$ and $\Var(Z_3 \mid X_2)$.
  \item Find $\E(Z_3 \mid \mathcal{F}_2)$ and $\Var(Z_3 \mid \mathcal{F}_2)$.
\end{enumerate}



\item Let $Y_t = W_t + 4t$ and consider the process $M_t = \exp(\alpha W_t - \alpha^2 t/2)$. 
The moment $\tau$ is the first moment when $Y_t$ hits $10$. 


\begin{enumerate}
  \item Check that $M_t$ is a martingale. 
  \item Find $f(t)$ such that $M_t = f(t) \exp(\alpha Y_t)$.
  \item Using Doob's theorem for $M_t$ find $\E(\exp(-(4\alpha +\alpha^2/2) \tau))$.
  \item Find $\E(\exp(-s \tau))$ for $s \geq 0$. 
  \item Find $\E(\tau)$.
\end{enumerate}
Hint: you may believe without penalty that Doob's theorem can be applied in this case. 

\end{enumerate}


\newpage
\section{HA-3}

\begin{enumerate}

\item Consider the process $dX_t = W_t^4 dW_t + W_t^6 dt$ with $X_0 = 2024$.
\begin{enumerate}
  \item Find $\E(X_t)$.
  \item Find $dY_t$ for $Y_t = X_t^2$.
  \item (bonus point) Find $\E(Y_t)$ and $\Var(X_t)$.
\end{enumerate}

\item Consider the process $C_t = W_t^3 + 2W_t^2 - 5 W_t^2 \cdot t\cdot W_t$.
\begin{enumerate}
  \item Find $dC_t$. 
  \item Is $C_t$ a martingale?
  \item Find the covariance $\Cov(C_t, \int W^2_u dW_u)$.
\end{enumerate}

\item Solve the stochatic differential equation
\[
dY_t = - Y_t dt + dW_t, \; Y_0 = 1
\]

If you are have no clues you may try a substitution $Z_t = f(t) Y_t$. 
Do not forget that the final answer may contain integrals that can't be calculated explicitely. It's ok.

\item Solve the stochatic differential equation
\[
dY_t = Y_t dt + (t^3 + 4Y_t) dW_t, \; Y_0 = 1
\]

If you are have no clues you may try to represent the process $Y_t$ as $Y_t = A_t B_t$, 
where $A_t$ is the solution of the equation $dA_t = A_t dt + 4A_t dW_t$. 

\item Consider the framework of Black and Scholes model: $S_t$ is the share price. 
Derive the current price of two European type assets, $X_0$ and $Y_0$. 

Future payoffs are given by:
\begin{enumerate}
  \item $X_T = (S_T - K)^3$ where $T$ and $K$ are fixed in the contract.
  \item $Y_T = S_T^{-2}$ where $T$ is fixed in the contract. 
\end{enumerate}

\item Consider the framework of Black and Scholes model with stochastic exchange rate. 
Now $S_t$ is the share price in dollars with $dS_t = \mu S_t dt + \sigma S_t dW_{1t}$.
The exchange rate $X_t$ (price of one dollar in euros) is driven by the equation 
$dX_t = \alpha X_t dt + \beta X_t dW_{2t}$. 

The Wiener processes $W_{1t}$ and $W_{2t}$ are independent, in particular that means $dW_{1t}dW_{2t} = 0$.

The risk free interest rate in euros is $r$.

The european type option pays you $\ln (S_t X_t)$ in euros at fixed time $T$. 

Find the price of this option in euros at time $t=0$, 
write your answer in terms of the share price in euros. 

\end{enumerate}



\end{document}







%\item Let's consider the process $X_t = \int_0^t u^2 dW_u$. 
%Prove that this process can be represented as a time changed Wiener process. 
%That means that there is a deterministic time-scaling $t(s)$ 
%such that $Y_s = X_{t(s)}$ is a Wiener process with respect to some filtration. 

